Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2024-02-26 main.tex
:pencil: Update remark 1

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	$\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{H}^{n \times n}$ and $\mathbb{U}^{n \times n}$ denotes the set of $n \times n$ Hermitian and unitary matrices, respectively.
	% $\mathbb{U}^{n \times n}$ denotes the set of $n \times n$ unitary matrices.
	$\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\Re\{\cdot\}$ takes the real part of a complex number.
	$\mathrm{tr}(\cdot)$ and $\det(\cdot)$ evaluates the trace and determinant of a square matrix, respectively.
	$\mathrm{diag}(\cdot)$ constructs a square matrix with arguments on the main diagonal and zeros elsewhere.
	$\mathrm{sv}(\cdot)$ returns the singular value vector.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ is the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^\dagger$ $(\cdot)^{(r)}$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), Moore-Penrose inverse, $r$-th iterated point, and final solution, respectively.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\lvert \cdot \rvert$ denotes the absolute value.
	% $\lVert \cdot \rVert _p$ means the $p$-norm and $\lVert \cdot \rVert$ suggests $p = 2$.
	$\lVert \cdot \rVert$ means the Euclidean norm.
	$\lVert \cdot \rVert _\mathrm{F}$ represents the Frobenius norm.
	$\mathcal{CN}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

@@

	Let $\mathbf{H}_\mathrm{D} \in \mathbb{C}^{N_\mathrm{R} \times N_\mathrm{T}}$, $\mathbf{H}_\mathrm{B} \in \mathbb{C}^{N_\mathrm{R} \times N_\mathrm{S}}$, $\mathbf{H}_\mathrm{F} \in \mathbb{C}^{N_\mathrm{S} \times N_\mathrm{T}}$ denote the direct (transmitter-receiver), backward (\gls{ris}-receiver), and forward (transmitter-\gls{ris}) channels, respectively.
	The equivalent channel is a function of the scattering matrix
	\begin{equation}
		\mathbf{H} = \mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} = \mathbf{H}_\mathrm{D} + \sum_g \mathbf{H}_{\mathrm{B},g} \mathbf{\Theta}_g \mathbf{H}_{\mathrm{F},g},
		\label{eq:channel_equivalent}
	\end{equation}
	where $\mathbf{H}_{\mathrm{B},g} \in \mathbb{C}^{N_\mathrm{R} \times L}$ and $\mathbf{H}_{\mathrm{F},g} \in \mathbb{C}^{L \times N_\mathrm{T}}$ are the backward and forward subchannels for \gls{ris} group $g$, corresponding to the $(g-1)L$ to $gL$ columns of $\mathbf{H}_\mathrm{B}$ and rows of $\mathbf{H}_\mathrm{F}$, respectively.
	% That is, $\mathbf{H}_{\mathrm{B},g}$ (resp. $\mathbf{H}_{\mathrm{F},g}$) corresponds to the $(g-1)L$ to $gL$ columns (resp. rows) of $\mathbf{H}_\mathrm{B}$ (resp. $\mathbf{H}_\mathrm{F}$).
	% , corresponding to the $(g-1)L$ to $gL$ columns of $\mathbf{H}_\mathrm{B}$ and rows of $\mathbf{H}_\mathrm{F}$.
	% $\mathbf{H}_{\mathrm{B},g}$ (resp. $\mathbf{H}_{\mathrm{F},g}$) corresponds to the $(g-1)L$ to $gL$ columns of $\mathbf{H}_\mathrm{B}$
	Since unitary matrices constitute a field with respect to multiplication, the scattering matrix of group $g$ can be decomposed as
	\begin{equation}
		\mathbf{\Theta}_g = \mathbf{L}_g \mathbf{R}_g^\mathsf{H},
	\end{equation}
	where $\mathbf{L}_g, \mathbf{R}_g \in \mathbb{U}^{L \times L}$ are two unitary factor matrices.
	Let $\mathbf{H}_{\mathrm{B},g} = \mathbf{U}_{\mathrm{B},g} \mathbf{\Sigma}_{\mathrm{B},g} \mathbf{V}_{\mathrm{B},g}^\mathsf{H}$ and $\mathbf{H}_{\mathrm{F},g} = \mathbf{U}_{\mathrm{F},g} \mathbf{\Sigma}_{\mathrm{F},g} \mathbf{V}_{\mathrm{F},g}^\mathsf{H}$ be the compact \gls{svd} of the backward and forward channels, respectively.
	The equivalent channel can thus be rewritten as
	\begin{equation}
		\mathbf{H} = \overbrace{\mathbf{H}_\mathrm{D} + \sum_g \mathbf{U}_{\mathrm{B},g} \mathbf{\Sigma}_{\mathrm{B},g} \underbrace{\mathbf{V}_{\mathrm{B},g}^\mathsf{H} \mathbf{L}_g \mathbf{R}_g^\mathsf{H} \mathbf{U}_{\mathrm{F},g}}_\text{backward-forward} \mathbf{\Sigma}_{\mathrm{F},g} \mathbf{V}_{\mathrm{F},g}^\mathsf{H}}^\text{direct-indirect}.
		\label{eq:channel_equivalent_svd}
	\end{equation}

	\begin{remark}
		By analyzing \eqref{eq:channel_equivalent_svd}, we conclude that channel shaping through \gls{bd}-\gls{ris} involves two critical challenges:
		\begin{itemize}
			% \item Subspace alignment: the singular matrices of backward and forward channels
			\item \textbf{Subspace alignment:} each group needs to balance the backward-forward (intra-group, multiplicative) and the direct-indirect (inter-group, additive) alignments of the singular matrices.
			\item \textbf{Subchannel rearrangement:} each group can rearrange and combine the backward and forward subchannels from strongest to weakest, which attains the maximal in rearrangement inequality.
		\end{itemize}
	\end{remark}

@@

	% \begin{remark}
	% 	From \eqref{eq:scattering_fc} and \eqref{eq:channel_equivalent_fc} in the proof of Proposition \ref{pp:fully_connected}, we notice that \eqref{iq:sv_bound_fc}--\eqref{iq:sv_bound_fc_power} are simultaneously tight when
	% 	% Interestingly, \eqref{iq:sv_bound_fc}--\eqref{iq:sv_bound_fc_power} are simultaneously tight when
	% 	\begin{equation}
	% 		\mathbf{\Theta} = \mathbf{V}_\mathrm{B} \mathbf{U}_\mathrm{F}^\mathsf{H}.
	% 		\label{eq:scattering_fc_tight}
	% 	\end{equation}
	% 	An interpretation is that the off-diagonal entries can enhance the capabilities of
	% 	\begin{itemize}
	% 		\item subspace alignment: $\mathbf{V}_\mathrm{B}$ and $\mathbf{U}_\mathrm{F}^\mathsf{H}$ in \eqref{eq:scattering_fc} fully align the subspaces of $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ by rotation;
	% 		\item subchannel rearrangement: $\mathbf{X} = \mathbf{I}$ in \eqref{eq:channel_equivalent_fc} pairs the subchannels of $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ from strongest to weakest, which attains the maximal in rearrangement inequality.
	% 	\end{itemize}
	% \end{remark}

>>>>

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	$\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{H}^{n \times n}$ and $\mathbb{U}^{n \times n}$ denotes the set of $n \times n$ Hermitian and unitary matrices, respectively.
	% $\mathbb{U}^{n \times n}$ denotes the set of $n \times n$ unitary matrices.
	$\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\Re\{\cdot\}$ takes the real part of a complex number.
	$\mathrm{arg}(\cdot)$ gives the argument of a complex number.
	$\mathrm{tr}(\cdot)$ and $\det(\cdot)$ evaluates the trace and determinant of a square matrix, respectively.
	$\mathrm{diag}(\cdot)$ constructs a square matrix with arguments on the main diagonal and zeros elsewhere.
	$\mathrm{sv}(\cdot)$ returns the singular value vector.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ is the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^\dagger$ $(\cdot)^{(r)}$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), Moore-Penrose inverse, $r$-th iterated point, and final solution, respectively.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\lvert \cdot \rvert$ denotes the absolute value.
	% $\lVert \cdot \rVert _p$ means the $p$-norm and $\lVert \cdot \rVert$ suggests $p = 2$.
	$\lVert \cdot \rVert$ means the Euclidean norm.
	$\lVert \cdot \rVert _\mathrm{F}$ represents the Frobenius norm.
	$\mathcal{CN}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

@@

	Let $\mathbf{H}_\mathrm{D} \in \mathbb{C}^{N_\mathrm{R} \times N_\mathrm{T}}$, $\mathbf{H}_\mathrm{B} \in \mathbb{C}^{N_\mathrm{R} \times N_\mathrm{S}}$, $\mathbf{H}_\mathrm{F} \in \mathbb{C}^{N_\mathrm{S} \times N_\mathrm{T}}$ denote the direct (transmitter-receiver), backward (\gls{ris}-receiver), and forward (transmitter-\gls{ris}) channels, respectively.
	The equivalent channel is a function of the scattering matrix
	\begin{equation}
		\mathbf{H} = \mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} = \mathbf{H}_\mathrm{D} + \sum_g \mathbf{H}_{\mathrm{B},g} \mathbf{\Theta}_g \mathbf{H}_{\mathrm{F},g},
		\label{eq:channel_equivalent}
	\end{equation}
	where $\mathbf{H}_{\mathrm{B},g} \in \mathbb{C}^{N_\mathrm{R} \times L}$ and $\mathbf{H}_{\mathrm{F},g} \in \mathbb{C}^{L \times N_\mathrm{T}}$ are the backward and forward subchannels for \gls{ris} group $g$, corresponding to the $(g-1)L$ to $gL$ columns of $\mathbf{H}_\mathrm{B}$ and rows of $\mathbf{H}_\mathrm{F}$, respectively.
	% That is, $\mathbf{H}_{\mathrm{B},g}$ (resp. $\mathbf{H}_{\mathrm{F},g}$) corresponds to the $(g-1)L$ to $gL$ columns (resp. rows) of $\mathbf{H}_\mathrm{B}$ (resp. $\mathbf{H}_\mathrm{F}$).
	% , corresponding to the $(g-1)L$ to $gL$ columns of $\mathbf{H}_\mathrm{B}$ and rows of $\mathbf{H}_\mathrm{F}$.
	% $\mathbf{H}_{\mathrm{B},g}$ (resp. $\mathbf{H}_{\mathrm{F},g}$) corresponds to the $(g-1)L$ to $gL$ columns of $\mathbf{H}_\mathrm{B}$
	Since unitary matrices constitute an algebraic group with respect to multiplication, the scattering matrix of group $g$ can be decomposed as
	\begin{equation}
		\mathbf{\Theta}_g = \mathbf{L}_g \mathbf{R}_g^\mathsf{H},
	\end{equation}
	where $\mathbf{L}_g, \mathbf{R}_g \in \mathbb{U}^{L \times L}$ are two unitary factor matrices.
	Let $\mathbf{H}_{\mathrm{B},g} = \mathbf{U}_{\mathrm{B},g} \mathbf{\Sigma}_{\mathrm{B},g} \mathbf{V}_{\mathrm{B},g}^\mathsf{H}$ and $\mathbf{H}_{\mathrm{F},g} = \mathbf{U}_{\mathrm{F},g} \mathbf{\Sigma}_{\mathrm{F},g} \mathbf{V}_{\mathrm{F},g}^\mathsf{H}$ be the compact \gls{svd} of the backward and forward channels, respectively.
	The equivalent channel can thus be rewritten as
	\begin{equation}
		\mathbf{H} = \overbrace{\mathbf{H}_\mathrm{D} + \sum_g \mathbf{U}_{\mathrm{B},g} \mathbf{\Sigma}_{\mathrm{B},g} \underbrace{\mathbf{V}_{\mathrm{B},g}^\mathsf{H} \mathbf{L}_g \mathbf{R}_g^\mathsf{H} \mathbf{U}_{\mathrm{F},g}}_\text{backward-forward} \mathbf{\Sigma}_{\mathrm{F},g} \mathbf{V}_{\mathrm{F},g}^\mathsf{H}}^\text{direct-indirect}.
		\label{eq:channel_equivalent_svd}
	\end{equation}

	\begin{remark}
		By analyzing \eqref{eq:channel_equivalent_svd}, we conclude that channel shaping through \gls{bd}-\gls{ris} creates two critical opportunities:
		\begin{itemize}
			\item \textbf{Subspace alignment:} each group can align the singular matrices of the backward-forward (intra-group, multiplicative) and the direct-indirect (inter-group, additive) channels. In \gls{siso}, subspace alignment boils down to phase matching and the optimal scattering matrix of group $g$ is
			\begin{equation}
				\mathbf{\Theta}_g^\star = \exp \bigl(\jmath \mathrm{arg}(h_\mathrm{D})\bigr) \mathbf{V}_{\mathrm{B},g} \mathbf{U}_{\mathrm{F},g}^\mathsf{H},
			\end{equation}
			where $\mathbf{V}_{\mathrm{B},g} = \bigl[\mathbf{h}_{\mathrm{B},g}/\lVert \mathbf{h}_{\mathrm{B},g} \rVert, \mathbf{N}_{\mathrm{B},g}\bigr]$, $\mathbf{U}_{\mathrm{F},g} = \bigl[\mathbf{h}_{\mathrm{F},g}/\lVert \mathbf{h}_{\mathrm{F},g} \rVert, \mathbf{N}_{\mathrm{F},g}\bigr]$, and $\mathbf{N}_{\mathrm{B},g}, \mathbf{N}_{\mathrm{F},g} \in \mathbb{C}^{L \times (L-1)}$ are the orthonormal bases of the null spaces of $\mathbf{h}_{\mathrm{B},g}$ and $\mathbf{h}_{\mathrm{F},g}$, respectively.
			Diagonal \gls{ris} is thus sufficient for perfect \gls{siso} phase matching (with empty null spaces), but its disadvantage in subspace alignment scales with \gls{mimo} dimensions, since each element can only apply a common phase shift to the ``pinhole'' indirect channel of size $N_\mathrm{R} \times N_\mathrm{T}$ passing through itself.
			Even if the \gls{bd}-\gls{ris} is fully-connected, there still exists a tradeoff between the direct-indirect and backward-forward alignments.
			% It suggests that diagonal \gls{ris} applying a phase shift to each indirect channel is sufficient for phase matching in \gls{siso}. but its disadvantage in subspace alignment scales with \gls{mimo} dimensions.
			% It suggests that diagonal \gls{ris} applying a phase shift to each indirect channel is sufficient for phase matching in \gls{siso}, but its disadvantage in subspace alignment scales with \gls{mimo} dimensions.
			% It is obvious that diagonal \gls{ris} is sufficient for perfect phase matching. However,
			% It is obvious that diagonal \gls{ris} is sufficient for perfect phase matching. However, in \gls{mimo} each element can only apply a common phase shift to the associated rank-1 $N_\mathrm{R} \times N_\mathrm{T}$ indirect channel.
			% Therefore, perfect subspace alignment of indirect channels through different elements is generally impossible.
			% It means the disadvantage of diagonal \gls{ris} in subspace alignment and subchannel rearrangement scales with \gls{mimo} dimensions.
			% is sufficient for perfect alignment
			% and is sufficient for perfect subspace alignment.
			% In \gls{mimo}, perfect subspace alignment is generally impossible due to the lack of direct control over the indirect channels.aa
			% backward-forward (intra-group, multiplicative) and the direct-indirect (inter-group, additive) alignments of the singular matrices.
			\item \textbf{Subchannel rearrangement:} this unique property of \gls{bd}-\gls{ris} allows each group to rearrange and combine the backward and forward subchannels by strength. In \gls{siso}, diagonal \gls{ris} provides a maximum indirect channel amplitude of $\sum_{n=1}^{N_\mathrm{S}} \lvert h_{\mathrm{B},n} \rvert \lvert h_{\mathrm{F},n} \rvert$, while \gls{bd}-\gls{ris} can generalize it to $\sum_{g=1}^{G} \sum_{l=1}^{L} \lvert h_{\mathrm{B},\pi_{\mathrm{B},g}(l)} \rvert \lvert h_{\mathrm{F},\pi_{\mathrm{F},g}(l)} \rvert$, where $\pi_{\mathrm{B},g}$ and $\pi_{\mathrm{F},g}$ are permutations of $\mathcal{L} \triangleq \{1, \ldots, L\}$. By rearrangement inequality, the maximum is attained by pairing the $l$-th strongest backward and forward subchannels in each group.
			The advantage of \gls{bd}-\gls{ris} in subchannel rearrangement scales with \gls{mimo} dimensions, since the number of subchannels associated with each group grows with $N_\mathrm{R} \times N_\mathrm{T}$.
			% effect is more pronounced in \gls{mimo} since the number of subchannels associated with each group grows quadratically with the dimensions.
			% The advantage of \gls{bd}-\gls{ris} in subchannel rearrangement scales with \gls{mimo} dimensions. since the number of subchannels associated with each group grows quadratically with the dimensions.
			% It suggests that the advantage of \gls{bd}-\gls{ris} in subchannel rearrangement scales with \gls{mimo} dimensions.
			% \item \textbf{Subchannel rearrangement:} this unique property of \gls{bd}-\gls{ris} allows each group to rearrange and combine the backward and forward subchannels by strength. In \gls{siso}, the maximum indirect channel amplitude by diagonal \gls{ris} is $\sum_{n=1}^{N_\mathrm{S}} \lvert h_{\mathrm{B},n} \rvert \lvert h_{\mathrm{F},n} \rvert$ while \gls{bd}-\gls{ris} can generalize it to $\sum_{g=1}^{G} \sum_{l=1}^{L} \lvert h_{\mathrm{B},\pi_{\mathrm{B},g}(l)} \rvert \lvert h_{\mathrm{F},\pi_{\mathrm{F},g}(l)} \rvert$, where $\pi_{\mathrm{B},g}$ and $\pi_{\mathrm{F},g}$ are independent permutations of $\mathcal{L} \triangleq \{1, \ldots, L\}$. By rearrangement inequality, the maximum is attained by pairing the $l$-th strongest backward and forward subchannels in each group.
			% \begin{equation}
			% 	\item \
			% \end{equation}
			% and $\sum_{g=1}^{G} \sum_{l=1}^{L} \lvert h_{\mathrm{B},\pi_{\mathrm{B},g}(l)} \rvert \lvert h_{\mathrm{F},\pi_{\mathrm{F},g}(l)} \rvert$ by \gls{bd}-\gls{ris}, where $\pi_{\mathrm{B},g}$ and $\pi_{\mathrm{F},g}$ are permutations of $\mathcal{L} \triangleq \{1, \ldots, L\}$. By rearrangement inequality,
			% \item \textbf{Subchannel rearrangement:} this unique property of \gls{bd}-\gls{ris} allows each group to rearrange and combine the backward and forward subchannels by strength. In \gls{siso}, the optimal indirect channel amplitude is $\sum_{n=1}^{N_\mathrm{S}} \lvert h_{\mathrm{B},n} \rvert \lvert h_{\mathrm{F},n} \rvert$ by diagonal \gls{ris} and $\sum_{n=1}^{N_\mathrm{S}} \lvert h_{\mathrm{B},\pi_\mathrm{B}(n)} \rvert \lvert h_{\mathrm{F},\pi_\mathrm{F}(n)} \rvert$ by \gls{bd}-\gls{ris}, where $\pi(\cdot)$ is a permutation of $\{1, \ldots, N_\mathrm{S}\}$.
			% this is unique to \gls{bd}-\gls{ris} --- each group can rearrange and combine the backward and forward subchannels by strength. In \gls{siso}, the optimal indirect channel amplitude is relaxed from $\sum_n \lvert h_{\mathrm{B},n} \rvert \lvert h_{\mathrm{F},n} \rvert$ to $\sum_n \lvert h_{\mathrm{B},\pi(n)} \rvert \lvert h_{\mathrm{F},\pi(n)} \rvert$ where $\pi(\cdot)$ is a permutation of $\{1, \ldots, L\}$.
			% from strongest to weakest, which attains the maximum in rearrangement inequality.
		\end{itemize}
	\end{remark}

@@

	% \begin{remark}
	% 	From \eqref{eq:scattering_fc} and \eqref{eq:channel_equivalent_fc} in the proof of Proposition \ref{pp:fully_connected}, we notice that \eqref{iq:sv_bound_fc}--\eqref{iq:sv_bound_fc_power} are simultaneously tight when
	% 	% Interestingly, \eqref{iq:sv_bound_fc}--\eqref{iq:sv_bound_fc_power} are simultaneously tight when
	% 	\begin{equation}
	% 		\mathbf{\Theta} = \mathbf{V}_\mathrm{B} \mathbf{U}_\mathrm{F}^\mathsf{H}.
	% 		\label{eq:scattering_fc_tight}
	% 	\end{equation}
	% 	An interpretation is that the off-diagonal entries can enhance the capabilities of
	% 	\begin{itemize}
	% 		\item subspace alignment: $\mathbf{V}_\mathrm{B}$ and $\mathbf{U}_\mathrm{F}^\mathsf{H}$ in \eqref{eq:scattering_fc} fully align the subspaces of $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ by rotation;
	% 		\item subchannel rearrangement: $\mathbf{X} = \mathbf{I}$ in \eqref{eq:channel_equivalent_fc} pairs the subchannels of $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ from strongest to weakest, which attains the maximum in rearrangement inequality.
	% 	\end{itemize}
	% \end{remark}
